Production scientific codes---weather, climate, and materials simulations---read
their behavior from \emph{configuration files} at startup and then never change it.
These configuration values, fixed for an entire multi-day run, are invisible to
optimizing compilers: they arrive through an I/O barrier into module-level state.
On a NAS~LU kernel we show this is not a missing optimization but a structural
property: an externally-linked configuration-style module global is not folded into
its loop bounds at any optimization level, because the value is absent from the program
and the global is not internalizable---and production configuration state is both.
\td{The strong form (that even a closed program with the value present folds it
\emph{cleanly}) needs the row-C redo with a source-constant injection; until then keep
the claim to the value-absence + linkage barriers, which are verified.} The result is a \emph{configuration tax}
paid on every timestep.
We present \emph{deployment-time specialization}: we statically recover the glacial
configuration variables a namelist fixes and inject them, so the standard
optimizer cascade---constant propagation, dead-code elimination, fusion,
vectorization---finally fires. Crucially, a recovered value is more than a scalar:
it also fixes \emph{types} and index \emph{ranges}. We therefore lift the recovered
domain under one \emph{interval-arithmetic-on-collections} lattice to a finite
type-set that devirtualizes polymorphic dispatch---our central mechanism, and one
the standard pipeline cannot perform even with the value in hand---and to integer
intervals that narrow index storage. Finite-domain configurations generate a
minimal covering set of pre-compiled, unguarded binaries dispatched once at
startup. We evaluate on the ICON climate model, ECMWF's ecRad radiation scheme, and
Quantum ESPRESSO across NVIDIA GH200 and AMD MI300A, comparing against
\texttt{nvfortran -O4} and against the strongest standard-tool baseline---the
recovered values injected as constants under whole-program LTO---which recovers the
scalar cases but, we expect, not the devirtualization or index narrowing.
\td{"devirt/narrowing carry the bulk of the speedup" is UNMEASURED (RQ4) --- state as a
hypothesis to evaluate, not a result; fill the magnitude once the eval lands.}
