\section{The Configuration Tax}
\label{sec:background}

\subsection{Configuration as glacial module state}
A production Earth-system or materials model is parameterized by a configuration file: a Fortran \emph{namelist} of a few hundred variables that select solvers, toggle physics, and fix grid and spectral dimensions. At startup the model executes \lstinline[style=fortran]|READ(NML=...)| once, depositing each value into a field of a derived type in a configuration module (in ICON, \texttt{mo\_run\_config}, \texttt{mo\_dynamics\_config}, \texttt{mo\_nonhydrostatic\_config}). Every consumer then reaches the value through \lstinline[style=fortran]|USE module, ONLY: var|. These variables are written once, at initialization, and only read during the thousands of timesteps that follow: they are \emph{glacial}~\cite{glacial}. They are also exactly the values a compiler cannot see: read behind an I/O barrier, held in module-level state, and consumed in a compute phase the compiler does not distinguish from initialization.

\subsection{Why \texttt{-O3} and LTO do not help}
\label{sec:whycompilers}
To test whole-program optimization precisely we take NAS~LU as a single translation unit whose \texttt{COMMON} blocks are rewritten to module variables, the same shape as a configuration module, and ask whether the LLVM pipeline can turn a representative grid bound (\texttt{nz}) into a compile-time constant in the loops that use it. \Cref{tab:lu} reports the outcome (flang-new-21, LLVM~21); a minimal microkernel and the full LU run reproduce every row.\footnote{Reproduction (HLFIR, LLVM IR at \texttt{-O0/-O2/-O3}, whole-program LTO, PGO, and a 22-way correctness harness): \texttt{experiments/lu\_globals/}.}

\begin{table}[t]
  \centering
  \caption{Making the LU grid bound \texttt{nz} a compile-time constant, measured on flang-new-21 / LLVM~21 (\texttt{experiments/lu\_globals}). When the value arrives at runtime through the namelist read, no level folds it, whole-program LTO included. Injecting the namelist variable as a source constant folds \texttt{nz0} but leaves the \emph{derived} bound \texttt{nz}\,$=$\,\texttt{nz0} non-constant; only rewriting the bound itself yields a literal. SCCP never participates: it is an SSA-value pass and globals are memory.}
  \label{tab:lu}
  \small
  \setlength{\tabcolsep}{4pt}
  \begin{tabular}{@{}p{5.7cm}l@{}}
    \toprule
    How \texttt{nz} is supplied & Loop bound \texttt{nz} \\
    \midrule
    runtime (namelist read), \texttt{-O3}               & runtime load \\
    runtime (namelist read), whole-program LTO          & runtime load \\
    namelist var \texttt{nz0} as source constant, \texttt{-O3} & runtime load \\
    namelist var \texttt{nz0} as source constant, LTO   & non-constant (\texttt{nz} guarded) \\
    loop bound rewritten to a source constant           & literal (folds) \\
    \bottomrule
  \end{tabular}
\end{table}

Two barriers keep the as-is bound a runtime value at every level, LTO included. \emph{Value presence}: with no constant in the program there is nothing to fold, because the value enters through the I/O barrier. \emph{Linkage}: an external-linkage module or \texttt{COMMON} global is not foldable in a single unit, since another unit may write it; only whole-program closure (LTO with internalization) removes this. A source \texttt{parameter} sidesteps both, which is why it folds at plain \texttt{-O3}.

A third barrier appears once the value is present. Inject the namelist variable \texttt{nz0} as a source constant and it folds, but the loops do not read \texttt{nz0}; they read \texttt{nz}, and \lstinline[style=fortran]|domain()| copies \texttt{nz = nz0} at runtime. That makes \texttt{nz} a write-once global. LLVM detects this (\texttt{GlobalStatus} marks it stored-once and \texttt{GlobalOpt} acts on it), but when a load may precede the store it cannot substitute the constant. It demotes \texttt{nz} to a one-bit ``initialized yet?'' flag and rematerializes the bound as \texttt{select i1 \%f, i32 16, i32 0}, a runtime guard rather than a constant. So even injection plus whole-program LTO, the strongest standard-tool baseline, leaves the trip count a runtime value; production code routinely launders configuration through such derived globals.

\subsection{What deployment-time specialization adds}
\Cref{tab:lu} delimits what is \emph{not} novel: once a scalar reaches a loop bound as a source constant with no derived indirection, the standard cascade (constant propagation, dead-code elimination, fusion) does the rest. Our contribution for that case is the bridge of \Cref{sec:recovery}, recovering the value statically from the configuration file and rewriting the bound to it, which is the one input the compiler structurally cannot obtain and the step that clears the derived-global guard. The contributions that go \emph{beyond} a supplied scalar, narrowing index storage from a recovered interval and devirtualizing from a recovered type-set (\Cref{sec:spec}), are optimizations no setting in \Cref{tab:lu} performs, and are where we expect the bulk of the speedup.

\td{DEEP ARG A (PLAN.md 2.A) --- phase-ordering / config-prop reparameterizes the
INPUT. Toolchain $B=(T_n\circ\dots\circ T_1)(S)$ ends in LTO; the config value
$v=F(\mathrm{var})$ is not derivable from $S$ at any phase, so the terminal LTO sink
cannot fold it. Config-prop is not a phase: it computes
$\mathrm{compile}(\mathrm{inject}(S,F))$. The injected-\texttt{nz0} rows are this with
inject done by hand (and still short of the derived bound); the as-is rows are
$\mathrm{compile}(S)$. "Why not LTO" fails on phase-ordering, not capability.}
\td{DEEP ARG D (PLAN.md 2.D) --- the "isn't this just recompile-per-config?"
three-layer floor: (1) UNAVAILABLE (values are runtime reads into mutable external
module state; baking in needs the glacial may-write proof = the contribution);
(2) INFEASIBLE (full per-config rebuild of a 1M-line model at every site; covering set
amortizes); (3) INSUFFICIENT (a scalar narrows no index type, devirtualizes no
dispatch). Consider a short Discussion subsection.}
