\section{Introduction}
\label{sec:intro}

Production scientific codes are launched, not merely run. Before it integrates a
single timestep, the ICON climate model~\cite{icon}, the ecRad radiation
scheme~\cite{ecrad2024}, or a Quantum ESPRESSO calculation reads a configuration
file---a Fortran \emph{namelist}---once, at startup, depositing a few hundred
values into module-level state that then never changes for the rest of a run that
lasts hours or days. These values select solvers, toggle physics, and fix grid and
spectral dimensions; having been written exactly once at initialization and read
throughout the compute phase, they are \emph{glacial}~\cite{glacial}. They are also
constants of the deployment in every sense but one: the compiler cannot see them.
The value lives in a data file the compiler never reads and, once loaded, in
externally-linked module state, so it remains a runtime value at every optimization
level. The result is a \emph{configuration tax}: branches, dead code, and fusion and
vectorization inhibitors that survive every optimization level even though every one
of them is decided the moment the model is launched.

We remove this tax by recovering the config-fixed values statically, narrowing each
to its domain, and specializing the code to the deployment. The reframing that makes
this practical is that no profiling is needed: profile-guided optimization
instruments a run to learn which values are stable in practice, but for
configuration variables \emph{the configuration file already is the profile}, and an
exact one. This puts specialization at a point in the compilation lifecycle distinct
from the usual two. Ahead-of-time compilation does not know the deployment values;
just-in-time compilation knows them but pays a runtime cost on every execution. We
perform \emph{deployment-time specialization}: a config-file-driven ahead-of-time
specialization with startup binary selection, which parses the namelist, traces each
variable across the I/O barrier to the glacial module state it initializes, recovers
its domain, and injects it---paying no runtime cost in the timestep loop at all.

\paragraph{Why the compiler cannot reach it.}
One might expect aggressive optimization to recover these constants. It cannot, for
two structural reasons rather than a missing pass, which we make precise in
\Cref{sec:background} on a NAS~LU kernel whose grid bounds are module variables (as
in any code built from \texttt{COMMON} blocks or configuration modules). First, the
value is \emph{not in the program at all}: it enters through an \emph{I/O barrier}, a
namelist read that lowers to opaque runtime calls past which constant propagation
must assume anything. Second, it is \emph{module-level state}---external by linkage,
reached across hundreds of translation units through
\lstinline[style=fortran]|USE ... ONLY|---so a single translation unit cannot even
assume it is unwritten. Production configuration state fails on both counts, so the
as-is config global stays a runtime value at every optimization level, including
whole-program LTO (\Cref{tab:lu}). For the scalar case, recovery is precisely the
missing input: once the value is supplied \emph{as a source-level constant}, the
standard cascade---constant propagation, dead-code elimination, fusion---does the
rest. The contribution there is not the cascade but the bridge, recovering statically
and soundly a value the compiler structurally cannot obtain.

\td{DEEP ARG B (PLAN.md 2.B) --- the config file is an EXACT, GUARD-FREE profile.
Value-knowledge spectrum: source-const(parse) | other-TU(LTO) | config/glacial
(deployment, EXACT) | PGO(post-run, probabilistic, NEEDS GUARD) | runtime. RIFS must
guard because values are merely frequent; we never guard because the namelist is
authoritative. Chain: exact $\Rightarrow$ no per-call guard $\Rightarrow$ unguarded
whole-program specialization $\Rightarrow$ covering set is the natural deployment unit
(ties C1/C2/C3). Rename the hook to "exact, guard-free profile".}
\td{DEEP ARG C (PLAN.md 2.C) --- automatic staging (dissolves "just partial
evaluation"). A configured code is a TWO-STAGE program; PE/SFAC need a hand-supplied
binding-time annotation, we DISCOVER the boundary automatically and EXECUTE stage 1 at
deployment-compile-time. CAVEAT (skeptical review): the boundary is NOT the textual
READ (glacial fixpoint reached later via config derivation / MPI broadcast / restart
override) --- define it as the last point dominating timestep-loop entries with empty
compute-phase may-write set (detail in \Cref{sec:recovery}). Elevates Autrey--Wolfe +
extends to types.}

\paragraph{A concrete tax.}
\Cref{lst:motivating} shows what the tax looks like in code: a loop nest from ICON's
dynamical core, lightly distilled. Three kernels---\texttt{A}, \texttt{B}, and
\texttt{C}---sweep the same column of edge values, reusing the horizontal and
tangential velocities \texttt{vn} and \texttt{vt}. A production compiler keeps them
as three separate loop nests: it cannot fuse \texttt{A} with \texttt{C} across the
opaque \lstinline[style=fortran]|IF (.NOT. lvn_only)| guard, and it cannot drop the
\lstinline[style=fortran]|IF (istep == 1)| test, because both \texttt{lvn\_only} and
\texttt{istep} are read once from the namelist at startup and never change. So on a
memory-bandwidth-bound code, the velocities are streamed from memory three times
where once would do. The moment the values are known the barriers dissolve: with
\texttt{istep}\,$=$\,2 the block is dead; with \texttt{istep}\,$=$\,1 and
\texttt{lvn\_only}\,$=$\,\texttt{.true.}, kernel \texttt{B} vanishes and \texttt{A}
and \texttt{C} fuse into one column sweep; with
\texttt{lvn\_only}\,$=$\,\texttt{.false.}, \texttt{B} becomes unconditional and all
three fuse. The information that unlocks every one of these optimizations is sitting,
in plain text, in a file on disk.

\begin{lstlisting}[style=fortran,float=t,caption={A configuration-dependent fusion
barrier in ICON's \texttt{velocity\_tendencies} (distilled). \texttt{istep} and
\texttt{lvn\_only} are read once from the namelist at startup and never change,
yet a compiler must treat them as unknown.},label={lst:motivating}]
! istep, lvn_only: read once from the namelist, then never change.
IF (istep == 1) THEN
  DO jk = 2, nlev                          ! kernel A
    vn_ie(je,jk)       = wgtfac_e*vn + (1-wgtfac_e)*vn_km1
    z_kin_hor_e(je,jk) = 0.5*(vn**2 + vt**2)
  END DO
  IF (.NOT. lvn_only) THEN                 ! <-- config-guarded fusion barrier
    DO jk = 2, nlev                        ! kernel B
      z_vt_ie(je,jk)   = wgtfac_e*vt + (1-wgtfac_e)*vt_km1
    END DO
  END IF
  DO jk = 2, nlev                          ! kernel C
    z_w_concorr_me(je,jk) = vn*ddxn_z + vt*ddxt_z
  END DO
END IF
\end{lstlisting}

\paragraph{Beyond scalars: interval arithmetic on collections.}
A recovered value, however, is rarely just a scalar, and the cases a standard
pipeline cannot finish even with the value in hand are the ones that matter most.
We treat four domain kinds uniformly under one lattice we call \emph{interval
arithmetic on collections}: a \emph{singleton} folds to a constant; a small
\emph{finite value-set} (a boolean, a few integer modes) drives specialization over
the cross-product of its elements; an \emph{integer interval}, recovered from the
shape an index array is used to address, narrows that array to the smallest integer
type that fits; and a finite \emph{type-set} collapses dynamic dispatch. The last is
our central mechanism. A configuration variable that selects one of several linear
solvers has a finite type-set; lifting the recovered value to that type-set replaces
the v-table with an \lstinline[style=fortran]|if|-chain that configuration constants
prune to the single deployed solver. Neither index narrowing nor devirtualization
follows from injecting a scalar and re-running the optimizer; indeed the default
Fortran pipeline (flang, nvfortran) does not devirtualize from a known dynamic type
at all (\Cref{sec:spec}). To our knowledge this is the first analysis to drive
devirtualization from a configuration-recovered value domain (\Cref{sec:related}).

\paragraph{Why this matters.}
Recovering these gains by hand is a recognized, expensive craft. ECMWF's recent
twelvefold acceleration of the ecRad radiation scheme~\cite{ecrad2024} decomposes
into a spectral factor and a separable, roughly threefold ``code optimization''
factor; hard-coding the spectral g-point count into build flags so that loop bounds
become constant---manual configuration propagation---is one of the hand optimizations
that contribute to that code-optimization factor. Such work costs hundreds of
person-hours, must be redone per configuration, and is fragile: a mismatched setting
silently corrupts or crashes the model. Deployment-time specialization automates it,
and because it knows the dimensions at compile time it also lets the GPU backend emit
tighter kernels---constant launch bounds, unrolled loops, narrowed index types---that
a runtime-valued parameter forecloses.

\td{DEEP ARG E (PLAN.md 2.E) --- systems value: production codes are generic for
MAINTAINABILITY/DISTRIBUTION, not because any single deployment needs generality
(every deployment runs a fixed config), so the genericity is pure runtime overhead
(the configuration tax). The real proposition is recovering specialized performance
WITHOUT sacrificing generic-source maintainability --- what the manual ecRad route
cannot do. Deeper than any single speedup number; state it here.}

\paragraph{Contributions.}
\begin{itemize}
  \item We characterize the \emph{configuration tax} and show, on a NAS~LU kernel,
  that the barrier is structural for two independent reasons---value-behind-I/O and
  external linkage---so no optimization level recovers production configuration state
  (\Cref{sec:background}). We position \emph{deployment-time specialization}
  (config-file-driven AOT specialization with startup binary selection) against AOT,
  JIT, and PGO.

  \item \textbf{Glacial recovery across the I/O barrier} (\Cref{sec:recovery}): a
  static analysis that traces namelist reads to the glacial module-level state they
  initialize and recovers it as deployment-time constants---the value the standard
  pipeline is missing and structurally cannot obtain.

  \item \textbf{Type-domain devirtualization} (\Cref{sec:spec}, our centerpiece):
  we lift the value/interval lattice to a finite type-set; to our knowledge the
  first analysis to drive devirtualization from a configuration-recovered value
  domain, eliminating dispatch that the default Fortran pipeline (flang, nvfortran)
  does not remove even when the dynamic type is known.

  \item \textbf{The variant economy} (\Cref{sec:variants}): finite-domain
  cross-product specialization with greedy minimal-covering-set selection over a
  distribution of operational namelists, deployed as multiple \emph{unguarded}
  whole-program binaries dispatched once at startup.

  \item \textbf{Evaluation} (\Cref{sec:eval}) on ICON, ecRad, and Quantum ESPRESSO
  across GH200 and MI300A, against \texttt{nvfortran -O4}\footnote{Current nvfortran
  (26.3) \emph{deprecates and ignores} \texttt{-Mipa}, so even \texttt{-O4} performs
  no interprocedural constant propagation across the namelist read.} and against the
  strongest standard-tool baseline---the recovered values injected as constants
  under whole-program LTO, which captures the scalar cases but not the
  devirtualization and index narrowing.
\end{itemize}

We build on the DaCe dataflow compiler and its Fortran frontend, which gives the
whole-program view our specializations need without depending on the maturity of
Flang/LLVM LTO on production-scale Fortran.
